Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-28 WSSSPE1_report/WSSSPE1_summary_report.tex
2 more small fixes

Most of the papers in this area focused on visibility of software in
scientific papers, since scientific papers are the most widely
accepted documentation of achievement in science. It was noted that
there appear to be no widely accepted standards on how the use of
software towards a paper ought to be mentioned, and that journals,
citation style guides and other guides to scientific conduct are vague
about how to describe software. To address this, papers advocated the
need for a fixed identifier for software, either directly through a
mechanism such as a Digital Object
Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper
written to document the software (and perhaps its creation), a
``software paper''~\cite{Chue_Hong_WSSSPE}. However, as was pointed
out during the panel discussion, one of the problems with papers as
the cited product is that their author list is fixed in time, which
discourages potential contributors who are not on the original author
list from designing incremental improvements as integration work
rather than separate (and hence possibly rewritten) software
products~\cite{howison_incentives_2013}.

Another approach is to reduce the difficulty of citing all software
underlying a research paper. For example, scientists often work with
software that itself wraps other software, leading to attribution
stacking that can make it non-obvious or even difficult to determine
what attributions would be appropriate. Knepley et
al.~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by
which the software itself, after it has run, provides the user with a
set of citations, according to the pieces of code actually
executed. They describe a prototype implementation whereby the
citations are embedded in libraries and reported along with the
results, via a command-line
interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty
that attempting to acknowledge the contributions of all pieces of
dependent code within a paper faces the difficulty of creating very
long citation lists, straining the analogy of code used to papers
cited. Katz approaches this issue by proposing a system of
``transitive credit,'' recording dependencies and relative
contributions outside particular papers, relieving authors from the
responsibility of acknowledging each and every dependency. Instead
authors would acknowledge the percentage contribution of the software
they used directly and an external system would then be able to
recursively allocate that credit to those who had provided
dependencies. Finally Priem and Piwowar argued that machine learning
techniques could examine the body of published literature and extract
mentions of software, coping with the multitude of informal ways in
which authors mention software they used~\cite{Priem_WSSSPE}.  A point
raised in the panel discussion was that instead of asking users to
improve their software citation practices, one can also ask how
software projects can better monitor the literature to improve their
ability to show impact. For example, the NanoHub project scans the
literature using keywords and names of known users to discover papers
that are likely to have used their software and platform, and assigns
graduate students to read each paper, highlighting mentions of
software use and sometimes following up with the authors to identify
stories for demonstrating impact.  A process for tracking
software-using publications with the goal of increasing impact
visibility is now described at \url{http://publications.wikia.com}.

>>>>

Most of the papers in this area focused on visibility of software in
scientific papers, since scientific papers are the most widely
accepted documentation of achievement in science. It was noted that
there appear to be no widely accepted standards on how the use of
software towards a paper ought to be mentioned, and that journals,
citation style guides and other guides to scientific conduct are vague
about how to describe software. To address this, papers advocated the
need for a fixed identifier for software, either directly through a
mechanism such as a Digital Object
Identifier~\cite{Katz2_WSSSPE,Knepley_WSSSPE} or via a published paper
written to document the software (and perhaps its creation), a
``software paper''~\cite{Chue_Hong_WSSSPE}. However, as was pointed
out during the panel discussion, one of the problems with papers as
the cited product is that their author list is fixed in time, which
discourages potential contributors who are not on the original author
list from designing incremental improvements as integration work
rather than separate (and hence possibly rewritten) software
products~\cite{howison_incentives_2013}.

Another approach is to reduce the difficulty of citing all software
underlying a research paper. For example, scientists often work with
software that itself wraps other software, leading to attribution
stacking that can make it non-obvious or even difficult to determine
what attributions would be appropriate. Knepley et
al.~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by
which the software itself, after it has run, provides the user with a
set of citations, according to the pieces of code actually
executed. They describe a prototype implementation whereby the
citations are embedded in libraries and reported along with the
results, via a command-line
interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty
that attempting to acknowledge the contributions of all pieces of
dependent code within a paper faces the difficulty of creating very
long citation lists, straining the analogy of code used to papers
cited. Katz approaches this issue by proposing a system of
``transitive credit,'' recording dependencies and relative
contributions outside particular papers, relieving authors from the
responsibility of acknowledging each and every dependency. Instead
authors would acknowledge the percentage contribution of the software
they used directly and an external system would then be able to
recursively allocate that credit to those who had provided
dependencies~\cite{Katz2_WSSSPE}. Finally Priem and Piwowar argued that machine learning
techniques could examine the body of published literature and extract
mentions of software, coping with the multitude of informal ways in
which authors mention software they used~\cite{Priem_WSSSPE}.  A point
raised in the panel discussion was that instead of asking users to
improve their software citation practices, one can also ask how
software projects can better monitor the literature to improve their
ability to show impact. For example, the NanoHub project scans the
literature using keywords and names of known users to discover papers
that are likely to have used their software and platform, and assigns
graduate students to read each paper, highlighting mentions of
software use and sometimes following up with the authors to identify
stories for demonstrating impact.  A process for tracking
software-using publications with the goal of increasing impact
visibility is now described at \url{http://publications.wikia.com}.
